Correcting for bias in high-throughput sequencing data

### 2011-06-21 seqbias-paper.tex
various corrections from markus korth

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated with protocol-specific sequence bias, influenced by PCR
amplification, and differing primer affinities and mixtures, for example.  The
result is decreased accuracy in many applications, such as de novo gene
annotation and isoform quantification in RNA-Seq, and peak calling in
ChIP-Seq.

@@

A significant example is the effect of transcript length on quantification.  As
noted by \citet{Oshlack2009} transcript length biases discovery of
differentially expression, since more reads are aligned to longer transcripts,
granting higher statistical significance in tests of differential expression.
Recently, some methods have been developed attempting to account for this effect
in gene set enrichment tests \citep{Young2010,Gao2011}, leading to more accurate
downstream analysis.

@@

This bias, though observed primarily in the 5' end of a read, is not resolved by
trimming the reads prior to mapping \citep{Hansen2010} (see also, Supplementary
Section 1), suggesting it is not a result of erroneous base calling, and that a
more sophisticated means of correction is needed.

@@

Li, et. al., \cite{Li2010} propose two models. The first is a Poisson linear
model, in which read counts across a transcript follow inhomogeneous Poisson
process. The read count at position $i$ within the transcript is Poisson
distributed with parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of
weights determined by the nucleotide at each position surrounding the read
start, in addition to a term capturing the abundance of the transcript.

@@

To estimate the bias $b_i$, we must make estimates of the background sequence
probability $\Pr[s_i]$ and the foreground sequence probability $\Pr[ s_i | m_i
]$, the latter being the probability of the sequence given a read being sampled
from its position. Estimating bias is therefore a problem of finding a model of
sequence probability that sufficiently complex to capture the common features of
the training data yet avoids overfitting.

@@

The methods proposed \citet{Hansen2010} and \cite{Roberts2011} also treat bias
correction as a problem of estimating foreground and background sequence
probabilities. They differ primarily in how these sequence probabilities are
estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
notation) directly, given training data consisting of long, highly-expressed
transcripts.

@@

In our model, we do not rely on constraining the set of networks (e.g. to
trees), and instead approximate the NP-Hard problem of determining the optimal
network structure using a fast hill climbing algorithm. Furthermore, we
train our model \emph{discriminatively}; only parameters that are deemed
informative in discriminating between foreground and background sequences are
included in the model. We thus seek to train a model that reduces bias, without
including uninformative parameters that would only increase variance.

@@

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to algorithm described by \citet{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood.
\begin{align*}
\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) \\
&=
\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
\end{align*}

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
\caption{
The network structures learned on each of the data sets are examined. Positions
are relative to the read start, which is labeled 0. Hollow circles indicate
position that were not included in the model, being deemed uninformative, give
the other positions and edges. The number of parameters of parameters needed to
specify each model is listed in parenthesis below.}
\label{fig:models}
\end{figure*}

@@

We applied our procedure (labeled ``BN'') as well as those as well as those of
\citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
are implemented in the R packages \texttt{mseq} and \texttt{Genominator},
respectively. The procedures were applied to four publicly available data
sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
an unpublished data sets of our own (Table \ref{tab:datasets}).

@@

Testing was performed by cross-validation. Each method was trained on data taken
from chromosomes 1--8 of the genome from which the reads were mapped (including
chromosomes 2a and 2b of the Chimpanzee genome). For evaluation, we drew a set
of long, highly expressed exons from the remaining chromosomes. In particular, for
each reference sequence, beginning with the set of exons annotated by Ensembl
release 60 \cite{Hubbard2009}, we removed any exons with known alternate splice
sites, then chose the top 1000 exons by read count, restricting ourselves those
at least 100nt long.

@@

\citet{Hansen2010} recommends using all of the reads to estimate heptamer
frequencies used by their model. The training procedure works by simple tallying
of frequencies. The implementation of this model in the Genominator package we
found to use a great deal of memory, and were unable to train with the volume of
data we wished, so we reimplemented the model and trained it on all of the reads
aligned to chromosomes 1-8.

@@

Plotting the nucleotide frequencies (Figure  \ref{fig:freqs}) we observe an
obvious bias. To quantify the non-uniformity observed in these plots, we use the

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as, symmetrized Kullback-Leibler (KL)
divergence \citep{Kullback1951}.
%Comparing only nucleotide frequencies may miss
%dinucleotide or higher order bias. Thus, for a fair comparison, we compute the
%KL divergence for $k$-mers, with $k \in \{ 1 \dots 6 \}$.
$$D_k( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
where the sum is over all $k$-mers.

To avoid the risk of a small number of reads with many duplicates dominating the
measure, we do not count duplicate reads when computing the KL divergence. Since we
explicitly chose exons with high coverage, not counting duplicates may not
fully capturing the effects of the bias correction methods. Thus, for each exon,
we compute the number of duplicates for each read, and count only reads in the
upper half. Under the assumption of uniform sampling, the set of reads falling
in the upper half should not depend on sequence, and we should expect the KL
divergence to be low.

@@

\begin{figure*}
\centerline{\includegraphics[width=1.00\textwidth]{fig/kl.eps}}
\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
frequencies expected under the assumption of uniform sampling from within exons.
A large divergence indicates significant bias. Plotted here is the divergence
from unadjusted reads counts as well as after adjusting reads counts using each
method.}
\label{fig:kl}
\end{figure*}

@@

\begin{figure}
\centerline{\includegraphics[width=0.3\textwidth]{fig/pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we measure a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive when uniformity is increased,
and negative when decreased. Boxes are plotted to mark the 25\%, 50\%, and 75\%
quantiles, with whiskers extending to 1.5 times the inter-quartile range (i.e.,
the span between tho 25\% and 75\% quantiles).
}

@@

\begin{table}
\processtable{
    The $R^2$ goodness of fit statistic measures increased uniformity in read
    coverage, after correcting for bias. Here the median $R^2$ across the test
    exons is listed for each method and simple. A positive $R^2$ indicates a
    better fit.\label{tab:pois}}{
    \begin{tabular}{lcccc}\toprule
          & BN    & MART  & GLM   & 7mer \\\midrule
Wetterbom & 0.174 & 0.016 & 0.066 & -0.079 \\
Katze     & 0.280 & 0.243 & 0.158 & 0.033 \\
Bullard   & 0.267 & 0.163 & 0.224 & 0.157 \\
Mortazavi & 0.240 & 0.210 & 0.197 & 0.091 \\
Trapnell  & 0.289 & 0.289 & 0.248 & 0.138
    \end{tabular}
}{}
\end{table}

@@

Our method generalizes this approach, attempting to overcome this problem by
using an estimation of sequence probability that is more robust to overfitting
and can account for bias beyond the initial heptamer. In all our tests, this
approach was at least as effective as those of \citet{Li2010}, despite not
requiring gene annotations, or manual selection of training examples.

We have withheld any direct comparison to the method described by
\citet{Roberts2011} and implemented in Cufflinks \citep{Trapnell2010}. Though
this method is superficially similar to our own, a proper comparison is
difficult, as the software can not be applied independently of estimating
transcript abundance in FPKM (fragments per exonic kilobase, per million mapped
reads) using Cufflinks. Fairness would dictate that competing methods be
substituted in FPKM estimation, or that a separate interface be written to the
Cufflinks bias correction method---both comparison requiring significant effort.

Though the Cufflinks method and our own both use graphical models to estimate
sequence probabilities, we make no restriction on the graph other than
acyclicity, and go to considerable effort to efficiently approximate the optimal
structure for each data set, rather than using a fixed structure, as in
Cufflinks. A ``one size fits all'' approach likely works quite well in many
cases, yet the observed specificity of the bias to protocol and platform argues
against it. For example, the structures learned by our method (Figure
\ref{fig:models}), are considerably different between those sequenced on an
Illumina platform versus an AB platform.

Training our model on more reads leads to more accurate estimation of bias, but
an increasingly long training time. For example, in our tests, fitting our model to
100,000 reads from the Mortazavi data, training time was approximately 45
minutes, running on one core of a 3Ghz Intel Xeon processor. However, limiting
the training to 25,000 reads leads to a model that is nearly as accurate while
requiring under 4 minutes to train. A full discussion of this trade-off is in
Supplementary Section 5.

Because we do not require annotations, ChIP-Seq, and other high-throughput
sequencing experiments may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one ChIP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated, however our method is still able effectively correct for the bias
that was observed (see, Supplementary Section 2).
Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in ChIP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into ChIP-Seq peak calling algorithms, and potentially improve
accuracy.

>>>>

\section{Motivation:}
Quantification of sequence abundance in RNA-Seq and ChiP-Seq experiments is
often conflated by protocol-specific sequence bias, influenced by PCR
amplification, or differing primer affinities and mixtures, for example.  The
result is decreased accuracy in many applications, such as de novo gene
annotation and isoform quantification in RNA-Seq, and peak calling in
ChIP-Seq.

@@

A significant example is the effect of transcript length on quantification.  As
noted by \citet{Oshlack2009} transcript length biases discovery of
differential expression, since more reads are aligned to longer transcripts,
granting higher statistical significance in tests of differential expression.
Recently, some methods have been developed attempting to account for this effect
in gene set enrichment tests \citep{Young2010,Gao2011}, leading to more accurate
downstream analysis.

@@

This bias, though observed primarily in the $5'$ end of a read, is not resolved by
trimming the reads prior to mapping \citep{Hansen2010} (see also, Supplementary
Section 1), suggesting it is not a result of erroneous base calling, and that a
more sophisticated means of correction is needed.

@@

\citet{Li2010} propose two models. The first is a Poisson linear
model, in which read counts across a transcript follow inhomogeneous Poisson
process. The read count at position $i$ within the transcript is Poisson
distributed with parameter $\lambda_i$, where, $\log(\lambda_i)$ is the sum of
weights determined by the nucleotide at each position surrounding the read
start, in addition to a term capturing the abundance of the transcript.

@@

To estimate the bias $b_i$, we must make estimates of the background sequence
probability $\Pr[s_i]$ and the foreground sequence probability $\Pr[ s_i | m_i
]$, the latter being the probability of the sequence given a read being sampled
from its position. Estimating bias is therefore a problem of finding a model of
sequence probability that is sufficiently complex to capture the common features
of the training data yet avoids overfitting.

@@

The methods proposed \citet{Hansen2010} and \cite{Roberts2011} also treat bias
correction as a problem of estimating foreground and background sequence
probabilities. They differ primarily in how these sequence probabilities are
estimated. \citet{Li2010} estimate reweighting coefficients ($b_i$, in our
notation) directly, given training data consisting of long, highly expressed
transcripts.

@@

In our model, we do not rely on constraining the set of networks (e.g. to
trees), and instead approximate the NP-Hard problem of determining the optimal
network structure using a fast hill-climbing algorithm. Furthermore, we
train our model \emph{discriminatively}; only parameters that are deemed
informative in discriminating between foreground and background sequences are
included in the model. We thus seek to train a model that reduces bias, without
including uninformative parameters that would only increase variance.

@@

To determine the structure and parameters of the Bayesian network, we use a
hill-climbing approach similar to the algorithm described by \citet{Grossman2004}.
The network structure is determined by greedily optimizing the conditional
log-likelihood.
\begin{align*}
\ell &= \sum_{i=1}^{n} \log \Pr( x_i | s_i ) \\
&=
\sum_{i=1}^{n} \log \frac{ \Pr(s_i | x_i) \Pr( x_i ) }{
\sum_{x \in \{0,1\}} \Pr( s_i | x ) \Pr(x) } \\
\end{align*}

@@

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{fig/models.eps}}
\caption{
The network structures learned on each of the data sets are examined. Positions
are relative to the read start, which is labeled 0. Hollow circles indicate
positions that were not included in the model, being deemed uninformative, given
the other positions and edges. The number of parameters needed to
specify each model is listed in parenthesis below.}
\label{fig:models}
\end{figure*}

@@

We applied our procedure (labeled ``BN'') as well as those of
\citet{Li2010} (``GLM'' and ``MART'') and \citet{Hansen2010} (``7mer''), which
are implemented in the R packages \texttt{mseq} and \texttt{Genominator},
respectively. The procedures were applied to four publicly available data
sets \citep{Bullard2010, Mortazavi2008, Trapnell2010, Wetterbom2010}, as well as
an unpublished data set of our own (Table \ref{tab:datasets}).

@@

Testing was performed by cross-validation. Each method was trained on data taken
from chromosomes 1--8 of the genome from which the reads were mapped (including
chromosomes 2a and 2b of the Chimpanzee genome). For evaluation, we drew a set
of long, highly expressed exons from the remaining chromosomes. In particular, for
each reference sequence, beginning with the set of exons annotated by Ensembl
release 60 \cite{Hubbard2009}, we removed any exons with known alternate splice
sites, then chose the top 1000 exons by read count, restricting ourselves those
at least 100 nt long.

@@

\citet{Hansen2010} recommends using all of the reads to estimate heptamer
frequencies used by their model. The training procedure works by simple tallying
of frequencies. The implementation of this model in the Genominator package
uses a great deal of memory, and we were unable to train with the volume of
data we wished, so we reimplemented the model and trained it on all of the reads
aligned to chromosomes 1--8.

@@

Plotting the nucleotide frequencies (Figure  \ref{fig:freqs}) we observe an
obvious bias. To quantify the non-uniformity observed in these plots, we use the
symmetrized Kullback-Leibler (KL) divergence \citep{Kullback1951}.

If $f_x$ is the background frequency of a $k$-mer $x$, and $f'_x$ the observed
frequency, the KL divergence is computed as
$$D_k( f, f' ) = \sum_{x} \left( f_x \log( f_x / f'_x ) + f'_x \log( f'_x / f_x) \right)$$
where the sum is over all $k$-mers.

To avoid the risk of a small number of reads with many duplicates dominating the
measure, we do not count duplicate reads when computing the KL divergence. Since we
explicitly chose exons with high coverage, not counting duplicates may not
fully capture the effects of the bias correction methods. Thus, for each exon,
we compute the number of duplicates of each read and count only reads in the
upper half. Under the assumption of uniform sampling, the set of reads falling
in the upper half should not depend on sequence, and we should expect the KL
divergence to be low.

@@

\begin{figure*}
\centerline{\includegraphics[width=1.00\textwidth]{fig/kl.eps}}
\caption{The Kullback-Leibler divergence compares the frequency of $k$-mers
(here, for $k = 1$ and $k = 4$) surrounding the starts of aligned reads to the
frequencies expected under the assumption of uniform sampling from within exons.
A large divergence indicates significant bias. Plotted here is the divergence
from unadjusted reads counts as well as after adjusting read counts using each
method.}
\label{fig:kl}
\end{figure*}

@@

\begin{figure}
\centerline{\includegraphics[width=0.3\textwidth]{fig/pois-boxplot.eps}}
\caption{For each of the 1000 test exons, we measure a pseudo-coefficients of
determination $R^2$, equivalent to the improvement in log-likelihood under the
bias corrected model. The statistic is positive when uniformity is increased,
and negative when decreased. Boxes are plotted to mark the 25\%, 50\%, and 75\%
quantiles, with whiskers extending to 1.5 times the inter-quartile range (i.e.,
the span between the 25\% and 75\% quantiles).
}

@@

\begin{table}
\processtable{
    The median $R^2$ goodness fo fit statistic across test exons.
\label{tab:pois}}{
    \begin{tabular}{lcccc}\toprule
          & BN    & MART  & GLM   & 7mer \\\midrule
Wetterbom & 0.174 & 0.016 & 0.066 & -0.079 \\
Katze     & 0.280 & 0.243 & 0.158 & 0.033 \\
Bullard   & 0.267 & 0.163 & 0.224 & 0.157 \\
Mortazavi & 0.240 & 0.210 & 0.197 & 0.091 \\
Trapnell  & 0.289 & 0.289 & 0.248 & 0.138
    \end{tabular}
}
{
    The $R^2$ statistic measures increased uniformity in read coverage, after
    correcting for bias.  Here the median $R^2$ across the test exons is listed
    for each method and sample. A positive $R^2$ indicates a better fit.
}
\end{table}

@@

Our method generalizes this approach, attempting to overcome this problem by
using an estimation of sequence probability that is more robust to overfitting
and can account for bias beyond the initial heptamer. In all our tests, this
approach was at least as effective as those of \citet{Li2010}, despite not
requiring gene annotations or manual selection of training examples.

We have withheld any direct comparison to the method described by
\citet{Roberts2011} and implemented in Cufflinks \citep{Trapnell2010}. Though
this method is superficially similar to our own, a proper comparison is
difficult, as the software cannot be applied independently of estimating
transcript abundance in FPKM (fragments per exonic kilobase, per million mapped
reads) using Cufflinks. Fairness would dictate that competing methods be
substituted in FPKM estimation, or that a separate interface be written to the
Cufflinks bias correction method---both comparisons requiring significant effort.

Though the Cufflinks method and our own both use graphical models to estimate
sequence probabilities, we make no restriction on the graph other than
acyclicity. We go to considerable effort to efficiently approximate the optimal
structure for each data set rather than using a fixed structure, as in
Cufflinks. A ``one size fits all'' approach likely works quite well in many
cases, yet the observed specificity of the bias to protocol and platform argues
against it. For example, the structures learned by our method (Figure
\ref{fig:models}) are considerably different between those sequenced on an
Illumina platform versus an AB platform.

Training our model on more reads leads to more accurate estimation of bias, but
an increasingly long training time. For example, in our tests, fitting our model to
100,000 reads from the Mortazavi data, training time was approximately 45
minutes, running on one core of a 3 Ghz Intel Xeon processor. However, limiting
the training to 25,000 reads leads to a model that is nearly as accurate while
requiring under 4 minutes to train. A full discussion of this trade-off is in
Supplementary Section 5.

Because we do not require annotations, ChIP-Seq, and other high-throughput
sequencing experiments, may also benefit from our model. In a preliminary
investigation, we found the sequence bias in one ChIP-Seq experiment
\cite{Cao2010} was less than that observed in any of the RNA-Seq data we
evaluated; however, our method is still able to effectively correct for the bias
that was observed (see, Supplementary Section 2).
Protocol differences, as we have seen,
can result in significant differences in observed nucleotide frequencies, so we
may not safely assert that bias in ChIP-Seq data is always low.  Given the weak
assumptions made by our model, our estimation of bias could easily be
incorporated into ChIP-Seq peak-calling algorithms, and potentially improve
accuracy.
